\mythesistitleNOTSYNPOSIS

### 2012-02-27 src/010chap1.tex
proof read TILL Sec 3.2.2

  An element $b$ \textbf{covers} $a$ provided that there exists no
  third element $x \in X$ \stt $a \posetr x \posetr b$.  A directed
  (or undirected, but with direction implied by horizontal positioning
  of nodes) graph that represents $\posetr$ in terms of its covering
  relation is called the \textbf{Hasse diagram} of the poset.
\end{definition}

\begin{definition}
  \label{def:poset}    
  Let $X$ be a partially ordered set with $\preccurlyeq$ being the
  partial order on $X$.

  \begin{enumerate}
  \hangindent \defindent
%  \item \textbf{hasse diagram}
  \item An element $X_m \in X$ is a
    \textbf{maximal upper bound} of $X$ if $\nexists X_q \in X$ such
    that $X_m \preccurlyeq X_q$.  A maximal upper bound on $X$ is
    denoted by $mub(X)$.

@@

\begin{definition}[Interval graph, Path graph]
  \label{def:pathgraph}
  Let $G$ be a graph.
  \begin{enumerate}
  \hangindent \defindent
  \item $G$ is an \textbf{interval graph} if there a set of intervals
    $\cI$ \stt $G$ is isomorphic to the intersection graph of
    $\cI$. \[G \cong \bI(\cI)\]
  \item $G$ is a \textbf{path graph} if there exists a (undirected
    unrooted) tree $T$ and a set of paths from $T$, $\cP$ \stt $G$ is
    isomorphic to the intersection graph of $\cP$. \[G \cong
    \bI(\cP)\] Path graphs are also known as \uvgraphs (Undirected
    tree where paths intersect in a Vertex) \cite{mw86}.
  \item $G$ is a \textbf{chordal graph} if it has no induced cycle of
    length greater than 3. Moreover, a \textbf{chordal graph} is
    characterized as an intersection graph of subtrees of a tree.
  \end{enumerate}
\end{definition}

The set of interval graphs is a subset of set of path graphs which is
a subset of set of chordal graphs. In fact, there is another class
between chordal graphs and path graphs called RDV-graphs (Rooted
Directed tree where paths intersect in a Vertex)  \cite{mw86} which is
defined exactly as a path graph except that the trees are rooted and
directed.

@@

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free on at least one of the partitions in
\cite{at72}. The first linear time algorithm for COP testing (COT) was
invented by \cite{bl76} using a data structure called \PQtrees.  Since
then several COT algorithms have been invented -- some of which
involved variations of \PQtrees \cite{mm96,wlh01,mcc04}, some involved
set theory and ICPIA \cite{wlh02,nsnrs09}, parallel COT
algorithms\cite{as95,bs03,ly91} and certifying algorithms\cite{mcc04}.

The construction of \PQtrees in \cite{bl76} draws on the close
relationship of matrices with \COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity.  PQR trees is a
generalized data structure based on PQ trees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build PQR
trees. \cite{wlh02} describes the simpler algorithm for COT. Hsu also
invented PC trees \cite{wlh01} which is claimed to be much easier to
implement. \cite{nsnrs09} describes a characterization of
consecutive-ones property solely based on the cardinality properties
of the set representations of the columns (rows); every column (row)
is equivalent to a set that has the row (column) indices of the rows
(columns) that have one entries in this column (row). This is
interesting and relevant, especially to this thesis because it
simplifies COT to a great degree.

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed in \cite{co98}.

@@

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, that finding a Tucker's
forbidden submatrix of at least $k$ rows is NP complete.

@@

The problem of interest in this thesis, namely, tree path labeling
problem, is a natural generalization of the interval assignment
problem or the COT problem. The problem is defined as follows -- given
a set system $\cF$ from a universe $U$ and a target tree $T$, does
there exist a bijection from $U$ to the vertices of $T$ such that each
set in the system maps to a path in $T$.  We refer to this as the
{\CFTPL} problem or simply {\em tree path labeling} problem for an
input set system and target tree pair -- $(\cF,T)$. The special case
of the target tree being a path, is the interval assignment problem.
We focus on generalizing the notion of an ICPIA \cite{nsnrs09} to
characterize feasible path assignments.  We show that for a given set
system $\cF$, a tree $T$, and an assignment of paths from $T$ to the
sets, there is a feasible\footnote{The notion of {\em feasibility} is
  formally defined in Section~\ref{ch:prelims}.}  bijection between
$U$ and $V(T)$ if and only if the intersection cardinalities among any
three sets (not necessarily distinct) is equal to that of the
corresponding paths assigned to them and the input passes a filtering
algorithm (described in this paper) successfully.  This algorithmic
characterization gives a natural data structure that stores all the
\remove{relevant} feasible bijections between $U$ and $V(T)$. This
reduces the search space for the solution considerably from the
universe of all possible bijections between $U$ and $V(T)$ to only
those bijections that maintain the characterization.  Further, the
filtering algorithm is also an efficient algorithm to test if a tree
path labeling\footnote{The terms {\em tree path labeling} and {\em
    tree path assignment} are, in informal language,
  synonyms. Formally, the former refers to the bijection $\cl: \cF
  \rightarrow \cP$. The latter refers to the set of ordered pairs
  $\{(S, P) \mid S \in \cF, P \in \cP\}$. $\cP$ is a set of paths on
  $T$.} is feasible.

@@

We see in Section~\ref{sec:motive} that pairwise intersection
cardinality preservation is necessary and sufficient for an interval
assignment to be feasible for a given hypergraph\footnote{A {\em
    hypergraph} is an alternate representation of a set system and
  will be used in this thesis.}~\footnote{See Section~\ref{ch:prelims}
  for the formal definition.} and thus is a characterization for COP
\cite{nsnrs09}. In our work we extend this characterization and find
that trio-wise intersection cardinality preservation makes a tree path
labeling\footnote{A {\em tree path labeling} $\cl$ is a bijection of
  paths from the target tree $T$ to the hyperedges in given hypergraph
  $\cF$.}~\footnotemark[4] (TPL) feasible, which is a generalization
of the COP problem. This problem is defined by \FTPL (See
Section~\ref{sec:myresearchintro} for problem definition). %as follows.

@@

We give a necessary and sufficient condition by way of {\em
  Intersection Cardinality Preservation Path Labeling} (ICPPL) and a
filtering algorithm for {\FTPL} to output in affirmative. ICPPL
captures the trio-wise cardinality property described
earlier\footnote{See Section~\ref{sec:feasible} for the definition of
  ICPPL.}. This characterization can be checked in polynomial time.  A
relevant consequence of this constructive procedure is that it is
sufficient to iteratively check if three-way intersection
cardinalities are preserved.  In other words, in each iteration, it is
sufficient to check if the intersection of any three hyperedges is of
the same cardinality as the intersection of the corresponding paths.
Thus this generalizes the well studied question of the feasible
interval assignment problem which is the special case when the target
tree $T$ is simply a path \cite{wlh02,nsnrs09}.

Aside from checking if a given TPL is feasible, we also solve the
problem of computing a feasible TPL for a given hypergraph and target
tree, if one exists. This problem is the {\CFTPL} problem (See
Section~\ref{sec:myresearchintro} for problem definition). %, is defined as follows.

@@

We present a polynomial time algorithm for {\CFTPL} when the target
tree $T$ belongs to a special class of trees called {\em \kstars} and
when the hyperedges in the hypergraph $\cF$ have at most $k+2$
vertices (\CFTPLKTREE -- See Section~\ref{sec:myresearchintro} for
problem definition). A couple of examples of {\kstars} are given in
Figure~\ref{fig:ksubstar}.

@@

In spite of this being a restricted case, we believe that our results
are of significant interest in understanding the nature of {\sc Graph
  Isomorphism} which is polynomial time solvable in interval graphs
while being hard on path graphs\cite{kklv10}. {\kstars} are a class of
trees which are in many ways very close to intervals or paths. Each
ray\footnote{The path from a leaf to the root, the vertex with highest
  degree, is called a {\em ray} of the \kstar.}~\footnotemark[4] are
independent except for the root\footnote{The vertex with maximum
  degree in a {\kstar} is called {\em root}.}~\footnotemark[4] and
hence can be considered as an independent interval till the root. Our
algorithm builds on this fact and uses the interval assignment
algorithm\cite{nsnrs09} up until ``reaching'' the root and then uses
the trio-wise intersection cardinality (the extra condition in ICPPL
that generalizes ICPIA) check to resolve the ambiguity about which ray
the algorithm should ``grow'' the solution into in the next iteration.

We also have an algorithm for solving {\CFTPL} with no restrictions on
the target tree or set size which runs in exponential time.  This
algorithm finds a path labeling from $T$ by decomposing the problem
into subproblems of finding path labeling of subsets of $\cF$ from
subtrees of $T$. Given the fact that binary matrices naturally
represent a set system (see Section~\ref{sec:motive}) and that the
{\em overlap} relation between the sets involved is an obvious
equivalence relation, $\cF$ quite naturally partitions into
equivalence classes known as {\em overlap components}. In the context
of COP, overlap components were used in \cite{wlh02} and
\cite{kklv10}. Moreover, \cite{nsnrs09} discovered that these
equivalence classes form a total order in the case of \COP. We extend this to TPL and find
that when $\cF$ is a path hypergraph\footnote{If there exists an FTPL
  for a hypergraph $\cF$, it is called a path hypergraph.}, the
classes can be partially ordered as an in-tree in polynomial
time. Once $\cF$ is ``broken'' into overlap components, one must
identify the subtree of $T$ that it needs to map to and this is the
hard part which is currently open to be solved in polynomial time.

>>>>

  An element $b$ \textbf{covers} $a$ provided that there exists no
  third element $x \in X$ \stt $a \posetr x \posetr b$.  A directed
  (or undirected, but with direction implied by vertical positioning
  of nodes) graph that represents $\posetr$ in terms of its covering
  relation is called the \textbf{Hasse diagram} of the poset.
\end{definition}

\begin{definition}
  \label{def:poset} 
  Let $X$ be a partially ordered set with $\posetr$ being the
  partial order on $X$.

  \begin{enumerate}
  \hangindent \defindent
%  \item \textbf{hasse diagram}
  \item An element $x_m \in X$ is a
    \textbf{maximal upper bound} of $X$ if $\nexists a \in X$ such
    that $x_m \posetr a$.  A maximal upper bound on $X$ is
    denoted by $mub(X)$.

@@

\begin{definition}[Interval graph, Path graph]
  \label{def:pathgraph}
  Let $G$ be a graph.
  \begin{enumerate}
  \hangindent \defindent
\item $G$ is an \textbf{interval graph} if there exists a set of
  intervals $\cI$ \stt $G$ is isomorphic to the intersection graph of
  $\cI$. \[G \cong \bI(\cI)\]
  \item $G$ is a \textbf{path graph} if there exists a (undirected
    unrooted) tree $T$ and a set of paths $\cP$ from $T$ \stt $G$ is
    isomorphic to the intersection graph of $\cP$. \[G \cong
    \bI(\cP)\] Path graphs are also known as \uvgraphs (Undirected
    tree where paths intersect in a Vertex) \cite{mw86}.
  \item $G$ is a \textbf{chordal graph} if it has no induced cycle of
    length greater than 3. Moreover, a chordal graph is characterized
    as an intersection graph of subtrees of a tree.
  \end{enumerate}
\end{definition}

The set of interval graphs is a subset of the set of path graphs which
is a subset of the set of chordal graphs. In fact, there is another
class between chordal graphs and path graphs called RDV-graphs (Rooted
Directed tree where paths intersect in a Vertex) \cite{mw86} which is
defined exactly as a path graph except that the trees are rooted and
directed. There are other variants of path graphs studied in the
literature with unrooted directed trees or where intersection of paths
are in edges rather than vertices (DV, UE, DE, RDE) \cite{gm03}.
% Recognizing clique graphs of directed edge path 
% graphs
% UV DuallyChordal Swarcter and Bornstein [10] 
% DV DuallyDV Prisner and Swarcter [8] 
% RDV DuallyRDV Prisner and Swarcter [8] 
% UE 
% DE DuallyDE This paper 
% RDE = RDV DuallyRDV Prisner and Swarcter [8]

@@

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free on at least one of the vertex partitions in
\cite{at72}. The first linear time algorithm for COP testing (COT) was
invented by \cite{bl76} using a data structure called \PQtrees.  Since
then several COT algorithms have been invented -- some of which
involved variations of \PQtrees \cite{mm96,wlh01,mcc04}, some involved
set theory and ICPIA \cite{wlh02,nsnrs09}, parallel COT
algorithms\cite{as95,bs03,ly91} and certifying algorithms\cite{mcc04}.

The construction of \PQtrees in \cite{bl76} draws on the close
relationship of matrices with \COP to interval graphs. A \PQtree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity.  \PQRtree is a
generalized data structure based on \PQtrees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build
\PQRtrees. \cite{wlh02} describes the simpler algorithm for COT. Hsu
also invented \PCtrees \cite{wlh01} which is claimed to be much easier
to implement. \cite{nsnrs09} describes a characterization of
consecutive-ones property solely based on the cardinality properties
of the set representations of the columns (rows); every column (row)
is equivalent to a set that has the row (column) indices of the rows
(columns) that have one entries in this column (row). This is
interesting and relevant, especially to this thesis because it
simplifies COT to a great degree.

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate even when it has no COP. This enables a user
to verify the algorithm's result even when the answer is
negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem addressed in
\cite{co98}.

@@

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that
the resulting matrix has COP. This problem is known to be NP complete
due to \cite{b75-phd}. \cite{v82} proves, using a reduction to the
longest path problem, that finding a Tucker's forbidden submatrix of
at least $k$ rows is NP complete.

@@

The problem of interest in this thesis, namely, tree path labeling
problem, is a natural generalization of the interval assignment
problem or the COT problem. The problem is defined as follows -- given
a set system $\cF$ from a universe $U$ and a target tree $T$, does
there exist a bijection from $U$ to the vertices of $T$ such that each
set in the system maps to a path in $T$.  We refer to this as the
{\CFTPL} problem or simply {\em tree path labeling} problem for an
input set system and target tree pair -- $(\cF,T)$. The special case
of the target tree being a path, is the interval assignment problem.
We focus on generalizing the notion of an ICPIA \cite{nsnrs09} to
characterize feasible path assignments.  We show that for a given set
system $\cF$, a tree $T$, and an assignment of paths from $T$ to the
sets, there is a feasible\footnote{The notion of {\em feasibility} is
  formally defined in Section~\ref{ch:prelims}.}  bijection between
$U$ and $V(T)$ if and only if the intersection cardinalities among any
three sets (not necessarily distinct) is equal to that of the
corresponding paths assigned to them and the input passes a filtering
algorithm (described in this paper) successfully.  This algorithmic
characterization gives a natural data structure that stores all the
feasible bijections between $U$ and $V(T)$. This reduces the search
space for the solution considerably from the universe of all possible
bijections between $U$ and $V(T)$ to only those bijections that
maintain the characterization.  Further, the filtering algorithm is
also an efficient algorithm to test if a tree path
labeling\footnote{The terms {\em tree path labeling} and {\em tree
    path assignment} are, in informal language, synonyms. Formally,
  the former refers to the bijection $\cl: \cF \rightarrow \cP$. The
  latter refers to the set of ordered pairs $\{(S, P) \mid S \in \cF,
  P \in \cP\}$. $\cP$ is a set of paths on $T$.} is feasible.

@@

We see in Section~\ref{sec:motive} that pairwise intersection
cardinality preservation is necessary and sufficient for an interval
assignment to be feasible for a given hypergraph\footnote{A {\em
    hypergraph} is an alternate representation of a set system and
  will be used in this thesis, in line with \cite{kklv10} (see
  Section~\ref{ch:prelims} for the formal definition).} and thus is a
characterization for COP \cite{nsnrs09}. In our work we extend this
characterization and find that trio-wise intersection cardinality
preservation makes a tree path labeling\footnote{A {\em tree path
    labeling} $\cl$ is a bijection of paths from the target tree $T$
  to the hyperedges in given hypergraph $\cF$ (see
  Section~\ref{ch:prelims} for the formal
  definition).}% ~\footnotemark[4]
(TPL) feasible, which is a generalization of the COP problem. This
problem is defined by \FTPL (see Section~\ref{sec:myresearchintro} for
problem definition). %as follows.

@@

We give a necessary and sufficient condition by way of {\em
  Intersection Cardinality Preservation Path Labeling} (ICPPL) and a
filtering algorithm for {\FTPL} to output in affirmative. ICPPL
captures the trio-wise cardinality property described
earlier (see Section~\ref{sec:feasible} for the definition of
  ICPPL.). This characterization can be checked in polynomial time.  A
relevant consequence of this constructive procedure is that it is
sufficient to iteratively check if three-way intersection
cardinalities are preserved.  In other words, in each iteration, it is
sufficient to check if the intersection of any three hyperedges is of
the same cardinality as the intersection of the corresponding paths.
Thus this generalizes the well studied question of the feasible
interval assignment problem which is the special case when the target
tree $T$ is simply a path \cite{wlh02,nsnrs09}.

Aside from checking if a given TPL is feasible, we also solve the
problem of computing a feasible TPL for a given hypergraph and target
tree, if one exists. This problem is the {\CFTPL} problem (see
Section~\ref{sec:myresearchintro} for problem definition). %, is defined as follows.

@@

We present a polynomial time algorithm for {\CFTPL} when the target
tree $T$ belongs to a special class of trees called {\em \kstars} and
when the hyperedges in the hypergraph $\cF$ have at most $k+2$
vertices (\CFTPLKTREE -- see Section~\ref{sec:myresearchintro} for
problem definition). A couple of examples of {\kstars} are given in
Figure~\ref{fig:ksubstar}.

@@

In spite of this being a restricted case, we believe that our results
are of significant interest in understanding the nature of {\sc Graph
  Isomorphism} which is polynomial time solvable in interval graphs
while being hard on path graphs\cite{kklv10}. {\kstars} are a class of
trees which are in many ways very close to intervals or paths. Each
ray\footnote{The path from a leaf to the root, the vertex with highest
  degree, is called a {\em ray} of the \kstar (see
  Section~\ref{ch:prelims} for the formal definition).} are
independent except for the root\footnote{The vertex with maximum
  degree in a {\kstar} is called {\em root} (see
  Section~\ref{ch:prelims} for the formal definition).} and hence can
be considered as an independent interval till the root. Our algorithm
builds on this fact and uses the interval assignment
algorithm\cite{nsnrs09} up until ``reaching'' the root and then uses
the trio-wise intersection cardinality (the extra condition in ICPPL
that generalizes ICPIA) check to resolve the ambiguity about which ray
the algorithm should ``grow'' the solution into in the next iteration.

We also have an algorithm for solving {\CFTPL} with no restrictions on
the target tree or set cardinality which runs in exponential time.
This algorithm finds a path labeling from $T$ by decomposing the
problem into subproblems of finding path labeling of subsets of $\cF$
from subtrees of $T$. Given the fact that binary matrices naturally
represent a set system (see Section~\ref{sec:motive}) and that the
{\em overlap} relation (see Section~\ref{ch:prelims} for the formal
definition) between the sets involved is an obvious equivalence
relation, $\cF$ quite naturally partitions into equivalence classes
known as {\em overlap components} (see Section~\ref{ch:prelims} for
the formal definition). In the context of COP, overlap components were
used in \cite{wlh02} and \cite{kklv10}. Moreover, \cite{nsnrs09}
discovered that these equivalence classes have a certain total order
in the case of \COP (see Section~\ref{sec:norestraint} for the formal
definition).  We extend this to TPL and find that when $\cF$ is a path
hypergraph\footnote{If there exists an FTPL for a hypergraph $\cF$, it
  is called a {\em path hypergraph}.}, the classes can be partially ordered
as an in-tree in polynomial time. Once $\cF$ is ``broken'' into
overlap components, one must identify the subtree of $T$ that it needs
to map to and this is the hard part which is currently open to be
solved in polynomial time.
